A Shell-like Model for General Purpose Programming

### 2018-12-10 doc/paper.tex
paper updates

@@

\paragraph{Not Everything Is a String.}
The first core issue that prevents most shells from being viable for large programs is that only strings are allowed as values.

Several projects, including bash itself, have attempted to add lists and hash-maps as standard objects. Unfortunately, while they allow these values to be stored in variables, most projects do not allow these data structures to be passed into or returned from functions normally.

The \verb/es/ shell (TODO: cite) attempts to extend the standard shell model with lambda functions with lexical scoping. However,

@@

\paragraph{Interruption.}
A core requirement of pipe-based languages is a semantic of process \emph{interruption}: automatic clean-up of processes that will no longer be used. Consider the following pseudo-code for a pipe-based language:

@@

\begin{verbatim}
(file-lines ?fname) = (
  handle = (open-file $fname) %%! close-file $fname
  until (=> eof? $handle) (=> read-until "\n" $handle)
)
\end{verbatim}

@@

(TODO show race condition with eager-interruption)

@@

(TODO explain guiding principle: \emph{A process is interrupted exactly when it can no longer be woken up} and the addendum \emph{without spawning new processes})

When a process is blocked on a synchronous channel, it will be woken up as soon as another process communicates on the other end.
Therefore the appropriate time for a process to be interrupted is if it is blocked on a channel that will never receive any more operations.
In traditional channel-based systems, this is difficult to detect, because a reference to a channel could be passed anywhere in the program as a standard value, and be interacted with in any way at any time.

(TODO therefore we add the constraint that processes must only interact with specifically registered channels)

\paragraph{Masking Interruptions.} (TODO: explain \verb/produce/ and \verb/consume/)

\paragraph{Closing Permanence.}
(TODO: explain why we do not reopen channels)

@@

@@

%# fn pipe(lhs, rhs, inp, out) {
%#   c = channel()
%#   go fn () { lhs(inp, c); };
%#   rhs(c, out)
%# }
%#
%# collect = collector()
%# pipe_collect(curry(range, 10), curry(fan, 2,
%  fn(el, inp, out) { write(el + 1, out); }
%), inp);
%#

>>>>

% TODO: discuss this
%# fn pipe(lhs, rhs, inp, out) {
%#   c = channel()
%#   go fn () { lhs(inp, c); };
%#   rhs(c, out)
%# }
%#
%# collect = collector()
%# pipe_collect(curry(range, 10), curry(fan, 2,
%  fn(el, inp, out) { write(el + 1, out); }
%), inp);
%#

@@

\paragraph{Not Everything Is a String.}
The first core issue that prevents most shells from being viable for large programs is that only strings are allowed as first-class values.

Several projects\cite{haahr}\cite{duff}, including bash itself, have attempted to add lists and hash-maps as standard objects. Unfortunately, while they allow these values to be stored in variables, most projects do not allow these data structures to be passed into or returned from functions, or importantly, passed through pipes normally.
Since values in these languages cannot be passed through pipes, they are relegated to a second class of value, one for which pipe-based composition is not available.

Thus if we are to allow for pipe-based composition to be the core composition method for large programs, we must allow rich values to be passed through channels.

@@

Luckily, given the decision to allow rich values to be passed through channels, a buffer is not as necessary for performance purposes, since a single write may contain an arbitrarily large amount of data - or for that matter a pointer.

\paragraph{Interruption.}
A core requirement of pipe-based languages is a semantic of process \emph{interruption}: automatic clean-up of processes that will no longer be used. Consider the following Magritte code:

@@

\begin{lstlisting}
(file-lines ?fname) = (
  handle = (open-file $fname) %%! close-file $fname
  until (=> eof? $handle) (=> read-until "\n" $handle)
)
\end{lstlisting}

@@

An advantage of lazy interruption is that a user can identify precisely which points in the code have the potential to be interrupted - those points which run a \verb/put/ or a \verb/get/. Whereas with eager-interruption semantics any point in the code may be interrupted, introducing the need for users to either mark critical uninterruptible sections or be very careful about setting correct compensations in the case of mutative algorithms.

On the other hand, lazy interruption has the disadvantage that a producing process must do enough work to produce one more value than will be consumed. If each value is relatively cheap to produce, this is not a problem, but in the case that the values are expensive to produce, this would result in doing a large amount of unnecessary work.

Overall, we have decided that the predictability of lazy interruption is worth the tradeoff for extra values, and we discuss some features that may mitigate this problem in \emph{Future Work}.

@@

Given that multiple processes may be reading from and writing to a channel, it is often the case that a communicating process will end when there are other processes still communicating over the channel. It would not be appropriate in this case to interrupt other processes attached to the channel, as they are still able to communicate.

We define our guiding principle for the appropriate time to interrupt a process as: \emph{A process is interrupted exactly when it can no longer be woken up}.
When a process is blocked on a synchronous channel, it will be woken up as soon as another process communicates on the other end.
Therefore the appropriate time for it to be interrupted is if it is blocked on a channel that will never receive any more operations.
In traditional channel-based systems, this is difficult to detect, because a reference to a channel could be passed anywhere in the program as a standard value, and be interacted with in any way at any time.

Luckily, in pipe-based languages, the arrangement of pipes is usually specified at process spawn time. This enables us to maintain a process register inside of each channel, so that we can decide when all readers or all writers have detached from a channel, at which time the channel closes and interrupts all blocked processes.

\paragraph{Closing Permanence.}
Though we close a channel when all processes on the read or write sides close, there is always the possibility of new processes spawning and attempting to connect to a channel. In such a case, some systems leave open the possibility of \emph{reopening} a channel for further use. In UNIX, for example, the named-pipes API allows for persistent channels (TODO: cite!) that, being merely files, can be reopened at any time.

However, this can lead to some unexpected races between a channel closing and a new process spawning. Consider the example:

\begin{lstlisting}
c = (make-channel)
& count-forever > $c
& take 10 < $c
put 10 > $c
\end{lstlisting}

This example, while contrived, represents an unavoidable race with first-class channels - whether the process running \verb/take 10/ closes the channel before or
after the \verb/put 10/ invocation opens the channel for writing. If we allow channel reopening, we will either block forever, or insert the number 10 into the stream, depending on which happens first. However if we do not allow channel reopening, we can say that, if a process initiates a read or write on a closed channel, it is immediately interrupted. In this case, both sides of the former race have the same termination behavior - the process is closed when \verb/take 10/ returns.

\paragraph{Masking Interruptions.}

A user will often wish to continue after a particular output channel has closed, and in this case will need a facility to mask exceptions for particular channels. Consider the following example, which uses reservoir-sampling to select a uniform random element from a stream:

\begin{lstlisting}
(sample) = (
  hold = (get)
  i = 1
  each (?v =>
    $i = (inc $i)
    # mutate the $hold variable with probability 1/$i
    prob $i 1 && ($hold = %v)
  )

  put $hold
)
\end{lstlisting}

In this example, the function \verb/each/ will consume the entire input stream and mutate a variable with a given probability. After the entire input stream is consumed, we wish to output the resulting value stored in the \verb/$hold/ variable.

However, with the semantics described above, the \verb/each/ function will loop until it is given an interrupt signal from the input channel closing, which will interrupt the entire process and not continue to the following line. In order to avoid this situation, the \verb/each/ function must somehow indicate that it intends to consume the entire stream, and handle the interrupt signal.

While we plan to explore more general mechanisms for error handling, for cases such as this it suffices to provide two builtins, \verb/produce/ and \verb/consume/, each of which takes a single function to loop forever until the standard output or standard input respectively is consumed, whereby control flow continues after the invocation. Using these functions, we might define \verb/each/ as:

\begin{lstlisting}
(each ?fn) = (consume (=> %fn (get)))
\end{lstlisting}

With this definition, the call to \verb/consume/ will mask the interruption, and control flow within the \verb/sample/ function will proceed as normal.

@@

The \verb/rc/ shell \cite{duff} from the Plan 9 operating system from Bell Labs introduces a list data type to shell, and cleans up some of the historical surprising quotation behavior of bash. However its list type is not nestable - nested lists are automatically flattened at creation time. Additionally, these lists cannot be passed through channels.

The \verb/es/ shell \cite{haahr} attempts to extend the standard shell model with lambda functions with lexical scoping. These shell functions are marshalled into strings when passed through pipes (whether to shell functions or external programs), but are still written byte-by-byte and thus are subject to interleaving in concurrent cases.

\bibliographystyle{ACM-Reference-Format}
\bibliography{paper}

@@
